Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-03 paper.qmd
Update of introduction

# Abstract
<!-- teaser --> The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and LALALALA.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

<!-- zooming in on problem -->
With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space, there is a need to re-evaluate the risks of
collision and casualties due to uncontrolled debris re-entries.
Overseeing authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment mega-constellations
of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life. Some of these will 
inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

@@

...

>>>>

# Abstract
<!-- teaser --> The recent decades observed an increasing number of space activities and associated launches. 
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

<!-- zooming in on problem -->
With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space, there is a need to re-evaluate the risks of
collision and casualties due to uncontrolled debris re-entries.
The increasing risk landscape poses challenges in terms of oversight and risk management.
There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

The increase in space operations does not exist in a vacuum. 
The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].

@@

Aviation safety risk management is a well-established domain in air transportation. 
Despite the increasing need for situational awareness of space objects and their re-entry, the classical risk avoidance paradigm revolves around restricting the airspace.
While the probability of a strike is low, it can result in catastrophic consequences. 
The growing number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.
